EDB 2.0 How Eagle Europeana project improved the Epigraphic Database Bari

### 2016-01-22 source/28/EAGLE2016_Rocco_engrev.tex
Paper 28 (Rocco) edited and added to the pdfs folder

\noindent The Epigraphic Database Bari (EDB) is an `old' database, it dates back to the late 1980s , when Carlo Carletti\footnote{
He was professor of Epigrafia e antichità cristiane in the former Department of Classical and Christian Studies of Bari
University.}, inspired by Jory's experience indexing CIL VI\footnote{\citet{_corpus_1974}. The six volumes of the
computerized KeyWord-In-Context index to all the approximately 40,000 texts collected in the sixth volume of the Corpus
Inscriptionum Latinarum (CIL) are the outcome of a trailblazing work of arrangement and organization of the
inscriptions in a database, even if limited to their textual part.}, started a project of digitization of the
inscriptions commissioned by Christians from Rome between the third and eighth centuries, collected and edited in the
27,688 lemmas of the ICVR\footnote{\emph{The Corpus of Inscriptiones christianae urbis Romae} started in 1922 by A. Silvagni,
was published, mostly, by A. Ferrua, later supported by D. Mazzoleni and C. Carletti, between 1956 and 1992. Pursuing
the work of G.B. de Rossi of the mid-1800s (IC), ICVR registers the inscriptions by Christians found in the suburban
area of Rome, sorted in topographic order by consular road, then by catacomb. Inscriptions found inside the urban walls
or the recently discovered suburban ones aren’t yet included in the ICVR volumes.}. The data were stored in a data
processing program, called ICVR, running under MS-DOS, later converted to a database for Microsoft Access. It was
originally intended for internal use only.

@@

In addition to the text, the data processing program recorded for each inscription\footnote{ The program didn’t allow
the use of Greek fonts; in order to add texts in the Greek alphabet it was necessary to type Latin equivalent letters
in the MS Word Symbol font, as you can see in Fig.~\ref{fig:1}}: bibliographic data (Progressivo = ICVR Number), origin
(Provenienza), type of support and technique of execution (Supp. e tecn.), function (Funzione), the presence or
absence of Christograms (\emph{Signa Christi}), and dating (Datazione). All this information was expressed by alphanumeric
codes of few characters, according to a limitation imposed by the program (Fig.~\ref{fig:1}).

\begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img002.jpg} 
\caption{The inscription ICVR V, 15147 as it appears now in EDB, with the acronyms expanded, the image and the text
displayed in Greek characters.}
\label{fig:2}
\end{figure}

Adding metadata to the text, even if in the minimalistic form of alphanumeric codes, accomplished the goal of describing
the epigraphic object in its widest sense as an inscribed artifact. This feature, in particular relating to
geographical information, was even more meaningful according to the characteristics of a large part of the inscriptions
recorded in EDB. In fact, the original pertinence to a monument/container (catacomb) or to a particular area of it,
allows, with reasonable certainty, to determine the patronage by a member of the Christian community and moreover to
determine the chronology, even without specific references inside the text\footnote{\citet{carletti_inscriptiones_1994}; \citet[viii-ix]{felle_introduzione_1997}.}. Likewise reporting the presence of Christological monograms allows us to identify them as explicit symbols
of Christian faith and as chronological indicators\footnote{ The monogram consisting of the first two letters of the
name \begin{otherlanguage*}{greek}Χριστός\end{otherlanguage*} can be considered the archetype of these signs. They first appear in inscriptions at the
beginning of the fourth century.}.

@@

It's even possible to define the relationship between the epigraph and the cited bibliography (printed or digital):
identity, when it is an edition; integration, if it's the edition of another fragment of the same inscription;
opistographic, if it's the edition of the inscription on the back side; reuse, if it's the edition of another
inscription on the same support; comment, if it's a study on a related topic.

@@

Data on the original context are therefore organized hierarchically in three related fields containing controlled lists.
After selecting the area of the suburb identified by the name of the consular road - or by the number of the Augustan
regio for urban inscriptions - it's possible to select the monument from a list: a catacomb or part of it, if it is a
large and multi-layered one; a church; a public building or an urban area. The third field allows access to a further
associated list where the position of the epigraph inside the monument can be selected. For the catacombs in
particular, it’s possible to use these fields to annotate the gallery or the cubicle, named with the alphanumeric code
used in the maps published in ICVR.

@@

Is worth noting that this set of associated fields refers to the original position of the inscriptions and not to the
place where they have been found, unless the two data coincide, that is in the case of inscriptions in situ or suo loco
adplicitae, in accordance with the definitions of the ICVR.

To complete information on spatial data, all cemeterial contexts have been georeferenced, so that clicking on their name
opens a new window in Google Map that shows the modern entrance to the cemetery, with the address and its coordinates
(Fig.~\ref{fig:6}).

 \begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img006.png} 
\caption{The original context maps.}
\label{fig:6}
\end{figure}

As a case of study, for the inscriptions in situ pertaining to the Domitilla catacomb (Via Ardeatina) - the only
cemeterial complex with a georeferenced plan of almost its entire extension\footnote{ START-Projekt: Die
Domitilla-Katakombe in Rome (Institut für Kulturgeschichte der Antike - Österreichische Akademie der Wissenschaften).
\citet{orlandi_case_2014}.} - clicking on the alphanumeric code\footnote{ The codes are made up of a majuscule letter
relating to a region of the catacomb \ and by another element, digit or minuscule letter, relating to a precise
internal position, respectively a gallery or a cubiculum. } associated to the precise position of the inscriptions,
gallery or cubiculum (F04, in Fig.~\ref{fig:6}), opens the plan of the specific area (regio) in which the inscription is found.
In every plan, the inscriptions preserved in situ are placed and marked by ICVR and EDB number (Fig.~\ref{fig:6})\footnote{The
plans are available on START-Projektwebsite: \url{www.oeaw.ac.at/antike/index.php\&id=431}, clicking on EDB number opens a
window with the EDB record. }.

@@

A survey of terminologies intended for description of epigraphic objects in the ICVR volumes has been carried out and
has generated lists of controlled terms for some of the fields (Fig.~\ref{fig:8}). \ Type of support, Executing technique and
Function vocabularies are aimed at classifying the specific and peculiar materials, methods and functions of the
inscriptions encoded in EDB, as the traditional epigraphic taxonomies do not totally adhere to their features.

@@

The proper text of the inscription is stored in an apposite field, following the Krummrey - Panciera
conventions\footnote{\citet{krummrey_criteri_1980}; \citet{Panciera1991}.}, with some adjustments to make it possible to describe
specific and peculiar issues of the inscriptions encoded in EDB. In particular the so-called
{\textquotedbl}aberrant{\textquotedbl} forms are not {\textquotedbl}normalized{\textquotedbl} to the
{\textquotedbl}standard{\textquotedbl} model, if they are recognized as grapho-phonetic outcomes of linguistic
modifications of Latin and Greek.

While the fidelity to what is written on the stone - or other type of support - respects and takes into due account the
evolution of Greek and Latin languages, it compromises the comprehension of the text and greatly complicates the
text-based search of terms. A standard query system, in fact, is not able to match a query with all the inscriptions
containing different spellings of a word. To resolve this issue each inscription is stored in its original form and in
a “lemmatized” form, where each term is actually replaced with its corresponding lemma, possibly by taking into account
its inflexed forms\footnote{ On the contrary, if the compiler recognizes aberrant forms as outcomes of misstatements
and material mistakes of the stonecutters, he transcribes them with the appropriate corrections, following the Krummrey
- Panciera conventions. \citet{felle_perspectives_2014} e \citet{orlandi_improving_2014}.}.

In EDB, as in every epigraphic database, the transcription of the text provides a real pre-edition, with systematic
expansion of abbreviations, hypothesis of integration, and, if possible, an update version of the text, according to
recent publications. It includes, moreover, the description of non-alphabetic signs - in double round brackets - such
as figurative elements (fishes, birds, anchors, etc.) and Christological monograms. Other fields record onomastic
notes, critical apparatus and textual comments.

@@

The latter of these can navigate in the descriptive section of the website (About EDB, People) and in the list of cited
Publications. They also have access to the entire database using two research masks: a Quick search allows the user to
search in only one of the following fields: identifier EDB, bibliographic data and text; an Advanced Search provides
the opportunity to explore the database through multiple search criteria variously combined.

@@

Like today's web search engines, EDB provides an advanced text search in Latin and Greek - an integrated tool
facilitates writing in the Greek alphabet\footnote{ Greek Inputter 2, developed by J. Naughton, allows the user to
write in Greek using his/her usual keyboard and to easily type various Greek diacritical marks
(\url{http://babel.mml.ox.ac.uk/ naughton / polytonic-greek-inputter.html}).} - allowing users to obtain different results
according to a default syntax, in the case of a search either for a single word or for a set of terms, in sequence or
not. Additionally, it's possible to choose whether or not to consider epigraphic diacritical marks, Greek accents and
spirits, and capitals. The textual search can be combined with other metadata related to bibliographic, geographic and
material data, or to function, reuse, language and date, expressed in a single year or in defined intervals.

@@

\item With the same aim, the controlled lists of Type of support, Executing technique and Function have been integrated in
the corresponding vocabularies of the EAGLE community. Among other benefits, such as alignment and relations between
databases, clicking on specific terms opens an EAGLE vocabulary window with a translation into various languages. This
feature is particularly useful in the case of EDB which, following ICVR, uses Latin for definitions, without modern
language translations.

@@

CIL = Corpus Inscriptionum Latinarum, \ Berolini 1863 ss.

CIG = Corpus Inscriptionum Grecarum, I-IV, Berolini 1828-1877

IC = Inscriptiones christiane urbis Romae septimo saeculo antiquiores, I-II, G.B. de Rossi (a c.), Romae 1857-1861;
Supplementum, G. Gatti (a c.), Romae 1915.

IG = Inscriptiones Graecae, Berolini 1873 ss.

ICVR = Inscriptiones christianae urbis Romae septimo saeculo antiquiores. Nova series, I-X, A. Silvagni, A. Ferrua, D.
Mazzoleni, C. Carletti (a c.), Romae - In Civitate Vaticana 1922 ss.

IGVR = Inscriptiones grecae urbis Romae, I-IV, L. Moretti (a c.), Roma 1968-1990.

@@

>>>>

\noindent The Epigraphic Database Bari (EDB) is an `old' database, it dates back to the late 1980s , when Carlo Carletti\footnote{
He was professor of \emph{Epigrafia e antichità cristiane} in the former \emph{Department of Classical and Christian Studies} of Bari
University.}, inspired by Jory's experience indexing CIL VI\footnote{\citet{_corpus_1974}. The six volumes of the
computerized KeyWord-In-Context index to all the approximately 40,000 texts collected in the sixth volume of the \emph{Corpus
Inscriptionum Latinarum} (CIL) are the outcome of a trailblazing work of arrangement and organization of the
inscriptions in a \emph{database}, even if limited to their textual part.}, started a project of digitization of the
inscriptions commissioned by Christians from Rome between the third and eighth centuries, collected and edited in the
27,688 lemmas of the ICVR\footnote{\emph{The Corpus of Inscriptiones christianae urbis Romae} started in 1922 by A. Silvagni,
was published, mostly, by A. Ferrua, later supported by D. Mazzoleni and C. Carletti, between 1956 and 1992. Pursuing
the work of G.B. de Rossi of the mid-1800s (IC), ICVR registers the inscriptions by Christians found in the suburban
area of Rome, sorted in topographic order by consular road, then by catacomb. Inscriptions found inside the urban walls
or the recently discovered suburban ones aren’t yet included in the ICVR volumes.}. The data were stored in a data
processing program, called ICVR, running under MS-DOS, later converted to a database for Microsoft Access. It was
originally intended for internal use only.

@@

In addition to the text, the data processing program recorded for each inscription\footnote{ The program didn’t allow
the use of Greek fonts; in order to add texts in the Greek alphabet it was necessary to type Latin equivalent letters
in the MS Word \emph{Symbol} font, as you can see in Fig.~\ref{fig:1}}: bibliographic data (Progressivo = ICVR Number), origin
(Provenienza), type of support and technique of execution (Supp. e tecn.), function (Funzione), the presence or
absence of Christograms (\emph{Signa Christi}), and dating (Datazione). All this information was expressed by alphanumeric
codes of few characters, according to a limitation imposed by the program (Fig.~\ref{fig:1}).

\begin{figure}[hbp]
\centering
 \includegraphics[scale=0.4]{EAGLE2016Roccoengrev-img002.jpg} 
\caption{The inscription ICVR V, 15147 as it appears now in EDB, with the acronyms expanded, the image and the text
displayed in Greek characters.}
\label{fig:2}
\end{figure}

Adding \emph{metadata} to the text, even if in the minimalistic form of alphanumeric codes, accomplished the goal of describing
the epigraphic object in its widest sense as an inscribed artifact. This feature, in particular relating to
geographical information, was even more meaningful according to the characteristics of a large part of the inscriptions
recorded in EDB. In fact, the original pertinence to a monument/container (catacomb) or to a particular area of it,
allows, with reasonable certainty, to determine the patronage by a member of the Christian community and moreover to
determine the chronology, even without specific references inside the text\footnote{\citet{carletti_inscriptiones_1994}; \citet[viii-ix]{felle_introduzione_1997}.}. Likewise reporting the presence of Christological monograms allows us to identify them as explicit symbols
of Christian faith and as chronological indicators\footnote{ The monogram consisting of the first two letters of the
name \begin{otherlanguage*}{greek}Χριστός\end{otherlanguage*} can be considered the archetype of these signs. They first appear in inscriptions at the
beginning of the fourth century.}.

@@

It's even possible to define the relationship between the epigraph and the cited bibliography (printed or digital):
\emph{identity}, when it is an edition; \emph{integration}, if it's the edition of another fragment of the same inscription;
\emph{opistographic}, if it's the edition of the inscription on the back side; reuse, if it's the edition of another
inscription on the same support; \emph{comment}, if it's a study on a related topic.

@@

Data on the \emph{original context} are therefore organized hierarchically in three related fields containing controlled lists.
After selecting the area of the suburb identified by the name of the consular road - or by the number of the Augustan
\emph{regio} for urban inscriptions - it's possible to select the monument from a list: a catacomb or part of it, if it is a
large and multi-layered one; a church; a public building or an urban area. The third field allows access to a further
associated list where the position of the epigraph inside the monument can be selected. For the catacombs in
particular, it’s possible to use these fields to annotate the gallery or the cubicle, named with the alphanumeric code
used in the maps published in ICVR.

@@

Is worth noting that this set of associated fields refers to the original position of the inscriptions and not to the
place where they have been found, unless the two data coincide, that is in the case of inscriptions \emph{in situ} or \emph{suo loco
adplicitae}, in accordance with the definitions of the ICVR.

To complete information on spatial data, all cemeterial contexts have been georeferenced, so that clicking on their name
opens a new window in \emph{Google Map} that shows the modern entrance to the cemetery, with the address and its coordinates
(Fig.~\ref{fig:6}).

 \begin{figure}[hbp]
\centering
 \includegraphics[width=\columnwidth]{EAGLE2016Roccoengrev-img006.png} 
\caption{The \emph{original context} maps.}
\label{fig:6}
\end{figure}

As a case of study, for the inscriptions \emph{in situ} pertaining to the Domitilla catacomb (Via Ardeatina) - the only
cemeterial complex with a georeferenced plan of almost its entire extension\footnote{\emph{START-Projekt: Die
Domitilla-Katakombe in Rome (Institut für Kulturgeschichte der Antike - Österreichische Akademie der Wissenschaften)}.
\citet{orlandi_case_2014}.} - clicking on the alphanumeric code\footnote{ The codes are made up of a majuscule letter
relating to a region of the catacomb \ and by another element, digit or minuscule letter, relating to a precise
internal position, respectively a gallery or a \emph{cubiculum}. } associated to the precise position of the inscriptions,
gallery or \emph{cubiculum} (F04, in Fig.~\ref{fig:6}), opens the plan of the specific area (\emph{regio}) in which the inscription is found.
In every plan, the inscriptions preserved in situ are placed and marked by ICVR and EDB number (Fig.~\ref{fig:6})\footnote{The
plans are available on START-Projektwebsite: \url{www.oeaw.ac.at/antike/index.php\&id=431}, clicking on EDB number opens a
window with the EDB record. }.

@@

A survey of terminologies intended for description of epigraphic objects in the ICVR volumes has been carried out and
has generated lists of controlled terms for some of the fields (Fig.~\ref{fig:8}). \ \emph{Type of support}, \emph{Executing technique} and
\emph{Function} vocabularies are aimed at classifying the specific and peculiar materials, methods and functions of the
inscriptions encoded in EDB, as the traditional epigraphic taxonomies do not totally adhere to their features.

@@

The proper text of the inscription is stored in an apposite field, following the Krummrey - Panciera
conventions\footnote{\citet{krummrey_criteri_1980}; \citet{Panciera1991}.}, with some adjustments to make it possible to describe
specific and peculiar issues of the inscriptions encoded in EDB. In particular the so-called
``aberrant'' forms are not ``normalized'' to the
``standard'' model, if they are recognized as grapho-phonetic outcomes of linguistic
modifications of Latin and Greek.

While the fidelity to what is written on the stone - or other type of support - respects and takes into due account the
evolution of Greek and Latin languages, it compromises the comprehension of the text and greatly complicates the
text-based search of terms. A standard query system, in fact, is not able to match a query with all the inscriptions
containing different spellings of a word. To resolve this issue each inscription is stored in its original form and in
a ``lemmatized'' form, where each term is actually replaced with its corresponding lemma, possibly by taking into account
its inflexed forms\footnote{ On the contrary, if the compiler recognizes aberrant forms as outcomes of misstatements
and material mistakes of the stonecutters, he transcribes them with the appropriate corrections, following the Krummrey
- Panciera conventions. \citet{felle_perspectives_2014} e \citet{orlandi_improving_2014}.}.

In EDB, as in every epigraphic database, the transcription of the text provides a real pre-edition, with systematic
expansion of abbreviations, hypothesis of integration, and, if possible, an update version of the text, according to
recent publications. It includes, moreover, the description of non-alphabetic signs - in double round brackets - such
as figurative elements (fishes, birds, anchors, etc.) and Christological monograms. Other fields record onomastic
notes, \emph{critical apparatus} and textual comments.

@@

The latter of these can navigate in the descriptive section of the website (About EDB, People) and in the list of cited
Publications. They also have access to the entire database using two research masks: a \emph{Quick search} allows the user to
search in only one of the following fields: identifier EDB, bibliographic data and text; an \emph{Advanced Search} provides
the opportunity to explore the database through multiple search criteria variously combined.

@@

Like today's web search engines, EDB provides an advanced text search in Latin and Greek - an integrated tool
facilitates writing in the Greek alphabet\footnote{ \emph{Greek Inputter 2}, developed by J. Naughton, allows the user to
write in Greek using his/her usual keyboard and to easily type various Greek diacritical marks
(\url{http://babel.mml.ox.ac.uk/ naughton / polytonic-greek-inputter.html}).} - allowing users to obtain different results
according to a default syntax, in the case of a search either for a single word or for a set of terms, in sequence or
not. Additionally, it's possible to choose whether or not to consider epigraphic diacritical marks, Greek accents and
spirits, and capitals. The textual search can be combined with other metadata related to bibliographic, geographic and
material data, or to function, reuse, language and date, expressed in a single year or in defined intervals.

@@

\item With the same aim, the controlled lists of \emph{Type of support}, \emph{Executing technique} and \emph{Function} have been integrated in
the corresponding vocabularies of the EAGLE community. Among other benefits, such as alignment and relations between
databases, clicking on specific terms opens an EAGLE vocabulary window with a translation into various languages. This
feature is particularly useful in the case of EDB which, following ICVR, uses Latin for definitions, without modern
language translations.

@@

\nocite{felle_fenomeni_2007,felle_esperienze_2012,jory_problems_1975}

@@

\section*{Corpora}
\begin{description}
\item[ CIL ] = \emph{Corpus Inscriptionum Latinarum}, \ Berolini 1863 ss.
\item[CIG ] = \emph{Corpus Inscriptionum Grecarum}, I-IV, Berolini 1828-1877
\item[IC] = \emph{Inscriptiones christiane urbis Romae septimo saeculo antiquiores}, I-II, G.B. de Rossi (a c.), Romae 1857-1861; \emph{Supplementum}, G. Gatti (a c.), Romae 1915. 
\item[IG] = \emph{Inscriptiones Graecae}, Berolini 1873 ss.
\item[ICVR] = \emph{Inscriptiones christianae urbis Romae septimo saeculo antiquiores. Nova series}, I-X, A. Silvagni, A. Ferrua, D. Mazzoleni, C. Carletti (a c.), Romae - In Civitate Vaticana 1922 ss.
\item[IGVR] = \emph{Inscriptiones grecae urbis Romae}, I-IV, L. Moretti (a c.), Roma 1968-1990.

\end{description}
